### 2021-09-26 documents/sections/proto_carib.tex
added additional split-s data, bakairi examples, rewrote up to sec 3.4

\begin{table}
	\centering
	\caption{Nominalizations of \gl{s_a_} and \gl{s_p_} verbs}
	\label{tab:nmlz}
	\begin{tabular}{@{}llll@{}}
	\mytoprule
Language & \gl{s_a_} verb & \gl{s_p_} verb & Source(s)\\
	\mymidrule
\kaxui & \obj{o-\emp{w-}ehurka-tpɨrɨ} \qu{your fall} & \obj{o-onenmehɨ-tpɨrɨ} \qu{your waking up} & \textcite[49, 74]{schuring2018kaxuyana}\\
\arara & \obj{\emp{w-}orik-tubo} \qu{dancing place} & \obj{ereŋmɨ-tpo} \qu{killing instrument} & \textcite[97]{alves2017arara} \\
\trio & \obj{ji-\emp{w-}əturu-to} \qu{(for) my talking} & \obj{j-emamina-to} \qu{(for) my playing} & \textcite[246]{triomeira1999} \\
\wayana & \obj{ɨ-\emp{w-}əturu-topo} \qu{my story} & \obj{j-ɨnɨkɨ-topo} \qu{my object for sleeping} & \textcite[409, 130]{wayanatavares2005}\\
\kalina & \obj{a-\emp{w-}ekupi-rɨ} \qu{your taking a bath} & \obj{aj-ereʔna-Ø} \qu{your fainting} & \textcite[392, 135]{hoff1968carib}\\
\apalai & \obj{j-epɨ-topo} \qu{my bath place} & \obj{j-enuru-topõ-pɨrɨ} \qu{the place of my birth} & \textcites[90]{koehn1986apalai}[ner2/1]{koehns1994textos}\\
\panare & \obj{j-\emp{u-}t͡ʃireema-n} \qu{their eating} & \obj{tj-arunkampətɨ-n} \qu{his hair standing on end} & \textcites[390]{panarepayne2013}[23]{mattei1994diccionario}\\
	\mybottomrule
	\end{tabular}
\end{table}

The distinction between \gl{s_a_} and \gl{s_p_} is also borne out in imperatives, the suffix for which can be reconstructed as \PC \rc{-kə}.
Here, \gl{s_p_} verbs took the \gl{p}-oriented second person prefix \rc{ə(j)-}, while \gl{s_a_} verbs were unprefixed.
This is illustrated with reflexes in various modern languages in \cref{tab:imp}.
As in the case of the \gl{s_a_} marker \rc{w-} participles and nominalizations, some languages have lost the distinction between \gl{s_a_} and \gl{s_p_} verbs in imperatives, for example \panare, or \kaxui (not shown in \cref{tab:imp}).

\begin{table}
	\centering
	\caption{Imperatives of \gl{s_a_} and \gl{s_p_} verbs}
	\label{tab:imp}
	\begin{tabular}{@{}llll@{}}
	\mytoprule
Language & \gl{s_a_} & \gl{s_p_} & Source(s)\\
	\mymidrule
\hixka & \obj{omoh-ko} \qu{come!} & \obj{\emp{oj-}okajɨm-ko} \qu{go up!} & \textcite[89, 44]{derbyshire1965textos}\\
\arara & \obj{odotpot-ko} \qu{come back!} & \obj{\emp{o-}alum-ko} \qu{jump!} & \textcite[161]{alves2017arara}\\
\trio & \obj{epɨ-kə} \qu{bathe!} & \obj{\emp{ə-}eremina-kə} \qu{sing!} & \textcite[323]{triomeira1999}\\
\wayana & \obj{əməm-kə} \qu{enter!} & \obj{\emp{əw-}eremi-kə} \qu{sing!} &  \textcite[227]{wayanatavares2005} \\
\kalina & \obj{oʔmaʔ-ko} \qu{stop!} & \obj{\emp{aj-}awon-ko} \qu{get up!} & \textcite[190]{hoff1968carib}\\
\apalai & \obj{otuʔ-ko} \qu{eat!} & \obj{\emp{o-}nɨʔ-ko} \qu{sleep!} & \textcites[62]{koehn1986apalai}[Mopo/20]{koehns1994textos} \\
\panare & \obj{ape-ʔ} \qu{flee!} & \obj{ahpən-kə} \qu{jump!} & \textcite[17, 5]{mattei1994diccionario}\\
	\mybottomrule
	\end{tabular}
\end{table}

@@

As for the form of this reflexive marker, \textcite[505--512]{meira2010origin} reconstruct two distinct  prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although they have since merged into a single morpheme in modern Cariban languages.
Reflexes of \detrz show diverse meanings, which are best subsumed under \dbqu{detransitive}; this range is illustrated with \trio examples in \exref{triodetrz}.

\ex<triodetrz> \trio \parencites[218--219]{meira2000split}[128, 256]{triomeira1999}\\
\begin{tabular}[t]{@{}llll@{}}
\\
\obj{nonta}  & → & \obj{e-nonta}, & \qu{abandon each other}\\
\qu{abandon} & & \obj{əi-nonta} &  (reciprocal) \\
\\
\obj{suka} & → & \obj{e-suka}, & \qu{wash self}\\
\qu{wash} & & \obj{əi-suka} & (reflexive)\\
\\
\obj{pahka} & → & \obj{e-pahka} & \qu{break (\gl{intr})}\\
\qu{break (\gl{tr})} & & & (anticausative)\\
\\
\obj{puunəpɨ} & → & \obj{əh-puunəpɨ}, & \qu{think, meditate}\\
\qu{think about} & & \obj{əi-puunəpɨ} & (antipassive)\\
\end{tabular}
\xe

%\pex<triodetrz> Semantics of the \trio detransitivizer \parencite[218--219]{meira2000split}
%\a \obj{e-suka}, \obj{əi-suka} \qu{to wash oneself} (reflexive)
%\a \obj{e-nonta}, \obj{əi-nonta} \qu{to abandon each other} (reciprocal)
%\a \obj{e-pahka} \qu{to break} (anticausative)
%\a \obj{əi-puunəəpɨ} \qu{to think} (antipassive)
%\xe
%
The morphological variation featured in \qu{to wash oneself} and \qu{to abandon each other} is due to the mentioned collapse between the two \PC prefixes:
\obj{e-} is a reflex of the reflexive prefix \rc{e-}, while the form \obj{əi-} originates in reciprocal \rc{əte-}.
However, both can occur with either meaning -- at least for these two verbs.

@@

\subsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
The original \PC split-\gl{s} system has been subject to change in many languages, in some even to the point of total loss.
Besides such modifications as for example losing the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} or losing certain person values altogether, these changes have largely been due to person prefixes being extended to new verbs.\footnote{
One could also imagine a scenario whereby verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
However, the verbs he sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial and explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system includes markers being extended to new verbs.}
There have been many such person marker extensions in Cariban languages, and some are still ongoing.
This is shown by \textcite{gildea1998}, using the Parukotoan languages as an example.
I have reproduced his tables as a tree diagram in \cref{fig:par_ext}, with adapted transcription -- and in the case of \kaxui, the addition of ∅/\obj{j-} as an alternative \gl{1}\gl{s_p_} marker \perscommpar{Spike Gildea}.
Apart from segmental changes to individual morphemes, the following restructuring innovations happened in the \setone paradigm in Parukotoan:

@@

While different cases of loss of split-\gl{s} are discussed by \textcite[91--96]{gildea1998}, this paper focuses on a so far neglected aspect of these person marking extensions.
I argue that they are executed via lexical diffusion, characterized as a type of extension by \textcite[106--115]{harris1995historical}; this hypothesis is supported by three facts.

@@

The remainder of this paper is structured as follows:
\cref{sec:extensions} investigates each of these six incomplete extensions individually, listing unaffected verbs and reconstructing proto-paradigms, where necessary.
\cref{sec:verbs} takes a comparative look at the verbs that were not affected by these extensions and searches for possible motivations.
Finally, \cref{sec:discussion} discusses the findings and puts them in a general context of language change and morphology.

>>>>

%\begin{table}
%	\centering
%	\caption{Nominalizations of \gl{s_a_} and \gl{s_p_} verbs}
%	\label{tab:nmlz}
%	\begin{tabular}{@{}llll@{}}
%	\mytoprule
%Language & \gl{s_a_} verb & \gl{s_p_} verb & Source(s)\\
%	\mymidrule
%\kaxui & \obj{o-\emp{w-}ehurka-tpɨrɨ} \qu{your fall} & \obj{o-onenmehɨ-tpɨrɨ} \qu{your waking up} & \textcite[49, 74]{schuring2018kaxuyana}\\
%\arara & \obj{\emp{w-}orik-tubo} \qu{dancing place} & \obj{ereŋmɨ-tpo} \qu{killing instrument} & \textcite[97]{alves2017arara} \\
%\trio & \obj{ji-\emp{w-}əturu-to} \qu{(for) my talking} & \obj{j-emamina-to} \qu{(for) my playing} & \textcite[246]{triomeira1999} \\
%\wayana & \obj{ɨ-\emp{w-}əturu-topo} \qu{my story} & \obj{j-ɨnɨkɨ-topo} \qu{my object for sleeping} & \textcite[409, 130]{wayanatavares2005}\\
%\kalina & \obj{a-\emp{w-}ekupi-rɨ} \qu{your taking a bath} & \obj{aj-ereʔna-Ø} \qu{your fainting} & \textcite[392, 135]{hoff1968carib}\\
%\apalai & \obj{j-epɨ-topo} \qu{my bath place} & \obj{j-enuru-topõ-pɨrɨ} \qu{the place of my birth} & \textcites[90]{koehn1986apalai}[ner2/1]{koehns1994textos}\\
%\panare & \obj{j-\emp{u-}t͡ʃireema-n} \qu{their eating} & \obj{tj-arunkampətɨ-n} \qu{his hair standing on end} & \textcites[390]{panarepayne2013}[23]{mattei1994diccionario}\\
%	\mybottomrule
%	\end{tabular}
%\end{table}

The distinction between \gl{s_a_} and \gl{s_p_} is also borne out in imperatives, the suffix for which can be reconstructed as \PC \rc{-kə}.
Here, \gl{s_p_} verbs took the \gl{p}-oriented second person prefix \rc{ə(j)-}, while \gl{s_a_} verbs were unprefixed.
This is illustrated with reflexes in various modern languages in \cref{tab:imp}.
As in the case of the \gl{s_a_} marker \rc{w-} participles and nominalizations, some languages have lost the distinction between \gl{s_a_} and \gl{s_p_} verbs in imperatives, for example \panare.

\begin{table}
	\centering
	\caption{Imperatives of \gl{s_a_} and \gl{s_p_} verbs \parencites[44, 89]{derbyshire1965textos}[161]{alves2017arara}[323]{triomeira1999}[227]{wayanatavares2005}[62]{koehn1986apalai}[Mopo/20]{koehns1994textos}[190]{hoff1968carib}[5, 17]{mattei1994diccionario}}
	\label{tab:imp}
	\input{floats/imperatives}
\end{table}

@@

As for the form of this reflexive marker, \textcite[505--512]{meira2010origin} reconstruct two distinct  prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although they have since merged into a single morpheme in modern Cariban languages.
Modern reflexes of \detrz show a range of meanings, which can all be characterized as \dbqu{detransitive}; this range is illustrated with \trio examples in \exref{triodetrz}.

\ex<triodetrz> \trio \parencites[218--219]{meira2000split}[128, 256]{triomeira1999}\\
\begin{tabular}[t]{@{}llll@{}}
\\
\obj{nonta}  & → & \obj{e-nonta}, & \qu{abandon each other}\\
\qu{abandon} & & \obj{əi-nonta} &  (reciprocal) \\
\\
\obj{suka} & → & \obj{e-suka}, & \qu{wash self}\\
\qu{wash} & & \obj{əi-suka} & (reflexive)\\
\\
\obj{pahka} & → & \obj{e-pahka} & \qu{break (\gl{intr})}\\
\qu{break (\gl{tr})} & & & (anticausative)\\
\\
\obj{puunəpɨ} & → & \obj{əh-puunəpɨ}, & \qu{think, meditate}\\
\qu{think about} & & \obj{əi-puunəpɨ} & (antipassive)\\
\end{tabular}
\xe
%
The morphological variation featured in \qu{to abandon each other} and \qu{to wash self} is due to the mentioned collapse between the two \PC prefixes:
\obj{e-} is a reflex of the reflexive prefix \rc{e-}, while the form \obj{əi-} originates in reciprocal \rc{əte-}.
However, both can occur with either meaning -- at least for these two verbs.

@@

\subsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
The original \PC split-\gl{s} system has been subject to change in many languages, in some even to the point of total loss.
Besides such modifications as for example losing the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} or losing certain person values altogether, these changes have largely been due to person prefixes being extended to new verbs.\footnote{
One could also imagine a scenario whereby verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
However, the verbs Meira sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial, thus explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system features markers being extended to new verbs.}
There have been many such person marker extensions in Cariban languages, and some are still ongoing.
This is shown by \textcite{gildea1998}, using the Parukotoan languages as an example.
We have reproduced his tables as a tree diagram in \cref{fig:par_ext}, with adapted transcription -- and in the case of \kaxui, the addition of ∅/\obj{j-} as an alternative \gl{1}\gl{s_p_} marker \perscommpar{Spike Gildea}.
Apart from segmental changes to individual morphemes, the following restructuring innovations happened in the \setone paradigm in Parukotoan:

@@

While different cases of loss of split-\gl{s} are discussed by \textcite[91--96]{gildea1998}, this paper focuses on a so far neglected aspect of these person marking extensions.
We argue that they are executed via lexical diffusion, characterized as a type of extension by \textcite[106--115]{harris1995historical}; this hypothesis is supported by three facts.

@@

The remainder of this paper is structured as follows:
In \cref{sec:extensions}, we discuss each of these six incomplete extensions individually, listing all attested unaffected verbs, and reconstructing proto-paradigms where necessary.
In \cref{sec:verbs}, we take a comparative look at the verbs that were not affected by these extensions and searches for possible motivations.
Finally, we discuss the findings and put them in a general context of language change and morphology in \cref{sec:discussion}.
